\documentclass[aps,prapplied,preprint,amsmath,amssymb,floatfix]{revtex4-2}
% If revtex4-2 is unavailable, switch the first line to:
%   \documentclass{article}
% and delete the \affiliation lines below.

\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}

\begin{document}

\title{When does weighted combination beat majority voting?\\
A six-study scoping rule, ending in a designed-to-kill test}

\author{Gavin Branaa}
\affiliation{Independent Researcher}

\date{July 5, 2026}

\begin{abstract}
Combining several unreliable binary judgments is a problem that recurs
across fault-tolerant engineering, neuroscience, machine-learning
ensembles, and everyday decision logic, and it admits two classical
answers: unweighted majority voting, and reliability-weighted (log-odds)
combination. Weighted combination is provably optimal when each source's
reliability is known; voting is the folk-robust default when it is not.
We report a six-study empirical program that sharpens this folklore into
a falsifiable scoping rule. Two studies against a real game's decision
logic first located both sides of the boundary: a vote-of-three beats an
information-poor OR-gate ($+1.8$\,pp, 30/30 seeds), and the same
vote-of-three loses decisively to an information-rich linear threshold
($-4.6$\,pp, 30/30 seeds) --- discretizing continuous evidence discards
margin information. Three independent-field checks then tested the rule
where each field's own theory makes a prediction: spiking-neuron
population decoding (weighted wins clean, voting wins by $+37$\,pp under
unmodeled per-trial outlier contamination), classical triple-modular
redundancy (weighted wins under stable calibration, the gap collapses to
noise under reliability drift), and scikit-learn ensembles under covariate
shift (weak, ambiguous --- the shift degraded accuracy without reordering
classifier reliability). The failure of the third check refined the rule
into its final form: \emph{weighted combination beats voting exactly as
long as the relative reliability ranking learned at calibration time
still holds at decision time; it degrades in proportion to rank
reordering, not to absolute miscalibration}. A final experiment was
designed to kill this refinement and failed to: a covariate shift
engineered (on validation data only) to flip the previously-best
classifier to worst reverses the weighted decoder's advantage (majority
wins, paired $t=-2.7$), while a damage-matched, rank-preserving shift
from the same shift family leaves weighted combination winning (contrast
$t=+2.6$), with a pooled dose--response between rank preservation and the
weighted advantage of $\rho=+0.44$ ($p=3.9\times10^{-7}$). All
experiments pre-registered their predictions in the code before running;
negative and weak results are reported alongside positive ones.
\end{abstract}

\maketitle

\section{Introduction}

Given $n$ unreliable binary judgments $v_i\in\{-1,+1\}$ of the same
latent truth, the two classical combiners are the unweighted majority
vote,
\begin{equation}
\hat y_{\mathrm{vote}}=\operatorname{sign}\Big(\sum_i v_i\Big),
\end{equation}
and the reliability-weighted (log-odds) combination,
\begin{equation}
\hat y_{\mathrm{wtd}}=\operatorname{sign}\Big(\sum_i w_i v_i\Big),\qquad
w_i=\log\frac{r_i}{1-r_i},
\label{eq:weighted}
\end{equation}
where $r_i$ is source $i$'s estimated reliability. When the $r_i$ are
known exactly and sources are independent, Eq.~\eqref{eq:weighted} is the
maximum-a-posteriori decoder and strictly dominates voting for
heterogeneous sources; this is textbook sensor fusion, and its ancestors
run from Condorcet's jury theorem~\cite{condorcet1785} through von
Neumann's multiplexed logic~\cite{vonneumann1956} to weighted-majority
online learning~\cite{littlestone1994} and classifier
combination~\cite{kittler1998,dietterich2000}. The folklore countervailing
principle, from robust statistics~\cite{huber1964}, is that when the
noise model is wrong, egalitarian estimators (votes, medians) resist
contamination that destroys weighted means.

Between ``weights are exactly right'' and ``weights are worthless'' lies
the entire practical regime: weights estimated from finite calibration
data and then deployed under drift, contamination, or distribution shift.
This paper reports an empirical program --- six pre-registered studies
across four unrelated substrates --- that locates the boundary and
compresses it into one falsifiable sentence:

\begin{quote}
\textbf{The scoping rule.} Weighted combination beats voting exactly as
long as the relative reliability \emph{ranking} learned at calibration
time still holds at decision time. It degrades in proportion to how much
the deployment regime \emph{reorders} relative reliability --- acute
per-trial reordering is worst, static reordering is measurable but
bounded by source diversity, and magnitude-only degradation, however
large, does not break it. Symmetrically: voting beats weighted
combination only where evidence is discretized or rankings are unstable;
where continuous margins are available and rankings hold, voting's
discretization is a pure loss.
\end{quote}

The program's origin is unglamorous and worth stating: the vote-of-three
primitive was first a hardware gate (a balanced-ternary CONSENSUS
instruction) and was tested against the live decision logic of a
tribe-simulation game. The point of this paper is not the game; it is
that the same rule then survived contact with three independent fields'
own established results, failed to be confirmed once in an instructive
way, and finally survived an experiment specifically constructed to
falsify its sharpest form.

\section{Studies 1--2: locating both sides of the boundary}

\paragraph{Study 1 (voting wins).} A game NPC's fight-or-flee decision
was implemented in the live code as \texttt{flee if (low\_hp OR
outnumbered)} --- logically a 2-input OR over discretized signals, with
no estimate of relative combat strength. Against a hidden ground-truth
combat model, adding a third noisy boolean (a rough strength estimate)
and taking a 2-of-3 vote beat the OR-gate by $+1.81$\,pp (0.7385 vs.\
0.7204; paired $t=71.7$; 30/30 seeds; 20\,000 trials/seed). Voting won
because the baseline was \emph{information-poor}: the vote added an
evidence channel and averaged discretization noise.

\paragraph{Study 2 (voting loses, predicted in advance).} The same
game's order-acceptance decision is already a weighted sum over
continuous evidence (\texttt{loyalty + courage} vs.\ a risk threshold).
Discretizing those same ingredients into three booleans and voting 2-of-3
lost by $-4.62$\,pp (0.8270 vs.\ 0.8732; $t=-107$; 30/30 seeds). The
loss was predicted in the test's docstring before running: votes discard
margin information a linear threshold already uses. Together, studies
1--2 fix the rule's first clause: the vote's value depends on what the
baseline already knows, not on any intrinsic virtue of consensus.

\section{Studies 3--5: three independent fields}

Each check pits decoder~\eqref{eq:weighted}, calibrated once on clean
data, against the unweighted vote, in a regime pair (calibration-valid
vs.\ calibration-violated), with the prediction stated before running:
weighted wins the first regime; voting wins or closes the gap in the
second.

\paragraph{Study 3: population coding (confirmed, dramatically).}
Fifteen leaky integrate-and-fire neurons encode a binary stimulus in
spike counts; an inverse-variance-weighted decoder and a per-neuron-vote
decoder are calibrated on clean trials. Clean regime: weighted wins,
0.9995 vs.\ 0.9958 ($t=6.4$, 25 seeds) --- real but near ceiling. Under
10\% per-trial unmodeled outlier contamination (an erratic spike burst on
a random subset of neurons each trial), the weighted decoder collapses to
0.610 while the vote holds 0.984 --- a $+37.4$\,pp voting win
($t=224$)~\footnote{One disclosed methodological step: the first clean-regime
run was at ceiling for both decoders; task difficulty was retuned before
final numbers. The contamination result was clear before and after.}.
This is the textbook robust-statistics mechanism made concrete: a single
large outlier swamps an inverse-variance-weighted sum, while the vote
needs only a majority of still-reliable neurons. Crucially, per-trial
contamination \emph{reorders} which neurons are trustworthy on each
individual trial --- no static weight can be right about that.

\paragraph{Study 4: triple-modular redundancy (confirmed, softly).} Five
independent binary units with heterogeneous reliabilities; weights
estimated from finite calibration observations. Stable regime: weighted
log-odds fusion wins, 0.9187 vs.\ 0.8912 ($+2.76$\,pp, $t=7.0$, 25
seeds). Under silent reliability drift ($\sigma=0.25$): the gap collapses
to statistical noise ($-0.35$\,pp, $t=-0.3$) --- it closes rather than
reverses. Drift only partially reorders the units' reliability ranking,
and the weighted decoder degrades proportionally --- a first hint that
reordering, not miscalibration per se, is the operative variable.

\paragraph{Study 5: ML ensembles under covariate shift (weak --- and the
refinement).} Five diverse scikit-learn classifiers (logistic regression,
decision tree, $k$NN, naive Bayes, small random forest), validation-set
log-odds weights, and a generic offset-and-rescale covariate shift on
40\% of features. The prediction \emph{did not clearly hold}: weighted
voting won the stable regime only marginally ($+0.72$\,pp, $t=2.1$, 20
seeds) and its edge did not collapse under shift ($+0.55$\,pp, $t=1.7$).
The honest diagnosis: the generic shift degraded all five classifiers'
absolute accuracy together without strongly reordering their relative
reliability, and a log-odds combiner mainly cares about ranking. This
weak result forced the rule into its final, sharper form (the ranking
clause above) --- and, read jointly with studies 3--4, predicted a
hierarchy: per-trial rank flips (population coding) $\gg$ partial drift
(TMR) $>$ rank-preserving shift (ensembles, no effect).

\section{Study 6: the designed-to-kill test}

The refined rule makes a specific counterfactual claim: it is the
\emph{reordering}, not the damage, that breaks the weighted decoder. Study
6 tests exactly that, on the study-5 substrate, with five regimes scored
on test labels that no selection step ever touches:

\begin{itemize}
\item[A.] no shift (replication control);
\item[B.] the original generic shift (replication of study 5);
\item[C.] a \emph{rank-reordering} shift: selected, from a fixed
mechanical candidate family (offset/rescale, sign-flip, column-permute on
random feature subsets), to make the previously-best validation
classifier maximally worse than its peers --- \emph{selection uses
validation data only};
\item[D.] a damage-matched generic control (intensity grid-searched on
validation to match C's mean accuracy damage);
\item[E.] the mirror control: drawn from the \emph{same} candidate family
as C, selected for maximum rank \emph{preservation} subject to damage
within tolerance of C's --- the exact mirror image of C's selection
rule~\footnote{Regime E was added in a disclosed second run after regime
D's manipulation check fell short (the generic family saturated at
13.4\,pp mean validation damage vs.\ C's 19.0\,pp). Regimes A--D and all
outcome measures were untouched and reproduce the first run exactly under
the same seeds.}.
\end{itemize}

\begin{table}[t]
\caption{Study 6: weighted-minus-majority accuracy gap across 30 seeds.
The manipulation check confirmed regime C flipped the previously-best
classifier into the bottom two at test time in 30/30 seeds; regimes C and
E have essentially identical absolute damage (19.3 vs.\ 18.9\,pp mean
accuracy drop) from the same shift family.}
\label{tab:study6}
\begin{ruledtabular}
\begin{tabular}{llrr}
Regime & Ranking & Gap (wtd$-$maj) & paired $t$ \\
\hline
A no shift          & preserved ($\rho=+0.92$) & $+0.85$\,pp & $+3.2$ \\
B generic shift     & preserved ($\rho=+0.74$) & $+0.59$\,pp & $+2.6$ \\
C rank-reordering   & \textbf{inverted} ($\rho=-0.39$) & $\mathbf{-0.30}$\,pp & $\mathbf{-2.7}$ \\
D damage-matched    & preserved ($\rho=+0.83$) & $+0.46$\,pp & $+2.8$ \\
E mirror control    & preserved ($\rho=+0.53$) & $+0.50$\,pp & $+2.0$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

Table~\ref{tab:study6} shows the result. The weighted decoder's advantage
survives $\sim$19\,pp of absolute damage when the ranking is preserved
(regime E) and reverses at the same damage when the ranking is inverted
(regime C); the paired C-vs-E contrast is $+0.80$\,pp ($t=+2.6$). Pooled
across all shifted regimes (120 seed$\times$regime points), the
weighted-minus-majority gap tracks the validation-to-test rank
correlation dose-responsively: Spearman $\rho=+0.44$,
$p=3.9\times10^{-7}$. All six pre-stated predictions held, including the
two replications.

The sign flip in regime C is statistically solid but small
($-0.3$\,pp), and this too is consistent with the rule: a static rank
flip mis-weights one voice among five diverse, clipped-weight voters ---
a bounded error --- whereas population coding's per-trial flips (study 3)
are unboundable by any static weight, which is why that collapse was
37\,pp. The effect-size hierarchy predicted by the refinement is exactly
the one observed across the program.

\section{Scope and limitations}

All six studies are simulations or synthetic-data experiments; no claim
is made about any particular production system. The decisions are binary
and sources conditionally independent by construction; correlated sources
favor neither combiner uniformly and are untested here. Study 6's regime
C is adversarial \emph{by design} --- the hypothesis under test is a
conditional (``if ranking reorders, then weighted breaks''), and how often
natural distribution shift is rank-reordering rather than magnitude-only
is an open empirical question this program does not answer. Weighted
combiners here use clipped log-odds of estimated accuracy; richer
combiners (stacking, temperature-scaled calibration) are out of scope.
Finally, the rule's symmetric clause (studies 1--2) concerns
\emph{discretization}: where continuous margins exist and rankings hold,
voting's loss is structural, not statistical.

\section{Practical reading}

For a designer, the rule reduces to two questions. (1)~\emph{Does my
baseline already use continuous margins?} If yes, do not replace it with
a vote (study 2); if it is a thresholded OR/AND over few signals, a
cheap vote over one more channel likely helps (study 1). (2)~\emph{Will
the reliability ranking I calibrated survive deployment?} If sources can
individually go rogue per decision (sensor glitches, per-trial outliers),
vote --- the weighted decoder's failure there is catastrophic (study 3).
If degradation is gradual and roughly shared (drift, generic shift),
calibrated weights keep a modest edge far longer than calibration
accuracy itself would suggest (studies 4--6), because ranking outlives
magnitude.

\section*{Methods and code}

Every study pre-registered its prediction in the experiment script's
docstring before the first run; first-run results are reported regardless
of outcome; the two disclosed deviations (study 3's difficulty retune,
study 6's added mirror control) are footnoted where they occur, and
neither touched an outcome measure after it was generated. Statistics are
paired $t$-tests across seeds (or Welch where unpaired), with seed counts
stated per study. All code, per-study findings documents, and raw outputs
are in the public repositories: studies 1--2 and 4--6 in
\texttt{github.com/gbranaa4-hue/012-trit-search} (\texttt{trit\_npc\_consensus\_test.py},
\texttt{trit\_order\_acceptance\_test.py}, \texttt{trit\_tmr\_test.py},
\texttt{trit\_ensemble\_test.py}, \texttt{trit\_rank\_reorder\_test.py});
study 3 in \texttt{github.com/gbranaa4-hue/Spikeling}
(\texttt{research/population\_coding\_test.py}).

% NOTE: verify bibliographic details before submission.
\begin{thebibliography}{9}
\bibitem{condorcet1785} M.~de Condorcet, \emph{Essai sur l'application de
l'analyse \`a la probabilit\'e des d\'ecisions rendues \`a la pluralit\'e
des voix} (Imprimerie Royale, Paris, 1785).
\bibitem{vonneumann1956} J.~von Neumann, ``Probabilistic logics and the
synthesis of reliable organisms from unreliable components,'' in
\emph{Automata Studies} (Princeton Univ.\ Press, 1956), pp.~43--98.
\bibitem{littlestone1994} N.~Littlestone and M.~K.~Warmuth, ``The
weighted majority algorithm,'' \emph{Inf.\ Comput.} \textbf{108}, 212
(1994).
\bibitem{kittler1998} J.~Kittler, M.~Hatef, R.~P.~W.~Duin, and J.~Matas,
``On combining classifiers,'' \emph{IEEE Trans.\ Pattern Anal.\ Mach.\
Intell.} \textbf{20}, 226 (1998).
\bibitem{dietterich2000} T.~G.~Dietterich, ``Ensemble methods in machine
learning,'' in \emph{Multiple Classifier Systems}, LNCS \textbf{1857}
(Springer, 2000), pp.~1--15.
\bibitem{huber1964} P.~J.~Huber, ``Robust estimation of a location
parameter,'' \emph{Ann.\ Math.\ Stat.} \textbf{35}, 73 (1964).
\bibitem{averbeck2006} B.~B.~Averbeck, P.~E.~Latham, and A.~Pouget,
``Neural correlations, population coding and computation,'' \emph{Nat.\
Rev.\ Neurosci.} \textbf{7}, 358 (2006).
\end{thebibliography}

\end{document}
